Verification of the CVM algorithm with a Functional Probabilistic Invariant

### 2025-06-28 paper/main.tex
fix author order
Paper: References to supplementary material.   - Instead of referring to directories in the anonymized zip file, we refer to the AFP entry and github repo.   - Improve related work section regarding distinct-elements algorithms from Bar-Yossed and Alon et al.   - Add license files (use BSD consistent with the license of the AFP entries.)   - Add readme for the CVM_Transform session which is going to be references by swhids.

% TODO: put in everyone's names here
\author{Emin Karayel}{Technical University of Munich, School of Computation, Information and Technology, Germany}{me@eminkarayel.de}{https://orcid.org/0000-0003-3290-5034}{}
\author{Seng Joe Watt}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{Watt_Seng_Joe@i2r.a-star.edu.sg}{https://orcid.org/0000-0002-6883-4736}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}
\author{Yong Kiam Tan}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore \and Nanyang Technological University, Singapore}{yongkiam.tan@ntu.edu.sg}{https://orcid.org/0000-0001-7033-2463}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}
\author{Kuldeep S. Meel}{Georgia Institute of Technology, United States \and  University of Toronto, Canada}{meel@cs.toronto.edu}{https://orcid.org/0000-0001-9423-5270}{Natural Sciences and Engineering Research Council of Canada (NSERC), funding reference [RGPIN-2024-05956]}
\author{Derek Khu}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{derek_khu@i2r.a-star.edu.sg}{https://orcid.org/0009-0000-0293-0664}{}

\authorrunning{E. Karayel, S. J. Watt, Y. K. Tan, K. S. Meel, D. Khu} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Emin Karayel, Seng Joe Watt, Yong Kiam Tan, Kuldeep S. Meel, Derek Khu} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

@@

\supplement{Isabelle/HOL Formalization}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%Template
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

% TODO Add Reference to transformation based proof.
\supplementdetails[subcategory={Isabelle/HOL Formalization}, cite=CVM_Distinct_Elements-AFP]{Software}{https://isa-afp.org/entries/CVM_Distinct_Elements.html}
\supplementdetails[subcategory={Isabelle/HOL Formalization}, cite=Negative_Association-AFP]{Software}{https://isa-afp.org/entries/Negative_Association.html}

>>>>

% TODO: put in everyone's names here
\author{Emin Karayel}{Technical University of Munich, School of Computation, Information and Technology, Germany}{me@eminkarayel.de}{https://orcid.org/0000-0003-3290-5034}{}
\author{Seng Joe Watt}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{Watt_Seng_Joe@i2r.a-star.edu.sg}{https://orcid.org/0000-0002-6883-4736}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}
\author{Derek Khu}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore}{derek_khu@i2r.a-star.edu.sg}{https://orcid.org/0009-0000-0293-0664}{}
\author{Kuldeep S. Meel}{Georgia Institute of Technology, United States \and  University of Toronto, Canada}{meel@cs.toronto.edu}{https://orcid.org/0000-0001-9423-5270}{Natural Sciences and Engineering Research Council of Canada (NSERC), funding reference [RGPIN-2024-05956]}
\author{Yong Kiam Tan}{Institute for Infocomm Research (I$^2$R), A*STAR, Singapore \and Nanyang Technological University, Singapore}{yongkiam.tan@ntu.edu.sg}{https://orcid.org/0000-0001-7033-2463}{Singapore NRF Fellowship Programme NRF-NRFF16-2024-0002}

\authorrunning{E. Karayel, S. J. Watt, D. Khu, K. S. Meel, Y. K. Tan} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Emin Karayel, Seng Joe Watt, Derek Khu, Kuldeep S. Meel, Yong Kiam Tan} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

@@

\supplement{Isabelle/HOL Formalization}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
\supplementdetails[cite=CVM_Distinct_Elements-AFP]{Software}{https://isa-afp.org/entries/CVM_Distinct_Elements.html}
\supplementdetails[cite=CVM_Transforms-Github, swhid=swh:1:dir:0d79e265cf5d4d19a0dd1d7e79ed0119cf39ae22]{Software}{https://github.com/joewatt95/CVM/tree/main/isabelle/CVM_Transforms}
\supplementdetails[cite=Negative_Association-AFP]{Software}{https://isa-afp.org/entries/Negative_Association.html}

### 2025-06-28 paper/sec/formalization.tex
Paper: References to supplementary material.   - Instead of referring to directories in the anonymized zip file, we refer to the AFP entry and github repo.   - Improve related work section regarding distinct-elements algorithms from Bar-Yossed and Alon et al.   - Add license files (use BSD consistent with the license of the AFP entries.)   - Add readme for the CVM_Transform session which is going to be references by swhids.

\section{Formalization of the CVM algorithm}\label{sec:formalization}
Let us now turn to details of our formalization of the CVM algorithm in Isabelle/HOL using our invariant-based approach.
We verified both the total, unbiased variant (\cref{alg:cvm_new}) and the original variant (\cref{alg:cvm}) from the introduction.

%TODO: make sure these names are correct!
\begin{note}
In our supplementary material, the invariant-based verification approach is in the directory \verb|CVM_Invariant|.
The theory \verb|CVM_Abstract_Algorithm| verifies a generalized version of the CVM algorithm, with an abstract subsampling operation that is required to fulfill \cref{i:subsample_condition}.
The specialization happens in the following theories, where \verb|CVM_Original_Algorithm| verifies the original algorithm, and \verb|CVM_New_Unbiased_Algorithm| verifies the new total and unbiased variant.
Note that only \verb|CVM_New_Unbiased_Algorithm| depends on the new library for negatively associated random variables, which we describe in more detail in \cref{sec:formalization_neg_dep}.
The total number of lines required for the verification of the original algorithm is \locnew~lines.
In addition, we actually verified a slight generalization of~\cref{alg:cvm} where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$; the original CVM algorithm~\cite{chakraborty2022} is the special case $f=\frac{1}{2}$.
\lipicsEnd\end{note}

>>>>

\section{Formalization of the CVM algorithm}\label{sec:formalization}
Let us now turn to details of our formalization of the CVM algorithm in Isabelle/HOL using our invariant-based approach~\cite{CVM_Distinct_Elements}.
We verified both the total, unbiased variant (\cref{alg:cvm_new}) and the original variant (\cref{alg:cvm}) from the introduction.

%TODO: make sure these names are correct!
\begin{note}
In our supplementary material~\cite{CVM_Distinct_Elements-AFP}, the theory \verb|CVM_Abstract_Algorithm| verifies a generalized version of the CVM algorithm, with an abstract subsampling operation that is required to fulfill \cref{i:subsample_condition}.
The specialization happens in the following theories, where \verb|CVM_Original_Algorithm| verifies the original algorithm, and \verb|CVM_New_Unbiased_Algorithm| verifies the new total and unbiased variant.
Note that only \verb|CVM_New_Unbiased_Algorithm| depends on the new library for negatively associated random variables, which we describe in more detail in \cref{sec:formalization_neg_dep}.
The total number of lines required for the verification of the original algorithm is \locnew~lines.
In addition, we actually verified a slight generalization of~\cref{alg:cvm} where the subsampling probability can be any $f \in [\frac{1}{2};e^{-1/12}]$; the original CVM algorithm~\cite{chakraborty2022} is the special case $f=\frac{1}{2}$.
\lipicsEnd\end{note}

### 2025-06-28 paper/sec/formalization_neg_dep.tex
fix spelling of Dubhashi
Paper: References to supplementary material.   - Instead of referring to directories in the anonymized zip file, we refer to the AFP entry and github repo.   - Improve related work section regarding distinct-elements algorithms from Bar-Yossed and Alon et al.   - Add license files (use BSD consistent with the license of the AFP entries.)   - Add readme for the CVM_Transform session which is going to be references by swhids.

\begin{note}
In our supplementary material, the formalization of negative association is in the directory \verb|Neg_Assoc|.
This library contains key results used to establish the invariants for CVM (e.g., \verb|Neg_Assoc_Permutation_Distributions|).
Although not needed for CVM, we have also mechanized the standard Chernoff bounds (\verb|Neg_Assoc_Chernoff_Bounds|), including the additive bounds by Hoeffding~\cite[Th. 1, 2]{hoeffding1963} and the multiplicative bounds by Motwani and Raghavan~\cite[Th. 4.1, 4.2]{motwani1995}.
Another example application included in the library is proving the false positive rate of Bloom filters (\verb|Neg_Assoc_Bloom_Filters|).
In total the library contains 2974~lines of Isabelle code.
\lipicsEnd\end{note}

@@

We relied on the proof by Dubashi et al.\ using the FKG inequality~\cite[Th. 10]{dubhashi1996}; there is a prior proof by Joag-Dev and Proschan~\cite[Th. 2.11]{joagdev1983}, which is incomplete\footnote{The step which we could not directly formalize is the assertion (Sentence~14) in the proof of Theorem~2.11 (\cite{joagdev1983}) that the conditional expectation of the random variable $f(X)$ is smaller iff the smallest element of the permutation is contained in $\mathrm{dep}(f)$. That statement is non-trivial and requires a proof using a theorem such as the FKG-inequality. We think Dubashi et al.\ developed their proof to complete it.}.
%However Dubashi presented a correct proof later using the FKG inequality~\cite[Th. 10]{dubhashi1996}.}

>>>>

\begin{note}
The formalization of the theory of negative association is included in a separate AFP entry~\cite{Negative_Association-AFP}.
This library contains key results used to establish the invariants for CVM (e.g., \verb|Neg_Assoc_Permutation_Distributions|).
Although not needed for CVM, we have also mechanized the standard Chernoff bounds (\verb|Neg_Assoc_Chernoff_Bounds|), including the additive bounds by Hoeffding~\cite[Th. 1, 2]{hoeffding1963} and the multiplicative bounds by Motwani and Raghavan~\cite[Th. 4.1, 4.2]{motwani1995}.
Another example application included in the library is proving the false positive rate of Bloom filters (\verb|Neg_Assoc_Bloom_Filters|).
In total the library contains 2974~lines of Isabelle code.
\lipicsEnd\end{note}

@@

We relied on the proof by Dubhashi et al.\ using the FKG inequality~\cite[Th. 10]{dubhashi1996}; there is a prior proof by Joag-Dev and Proschan~\cite[Th. 2.11]{joagdev1983}, which is incomplete\footnote{The step which we could not directly formalize is the assertion (Sentence~14) in the proof of Theorem~2.11 (\cite{joagdev1983}) that the conditional expectation of the random variable $f(X)$ is smaller iff the smallest element of the permutation is contained in $\mathrm{dep}(f)$. That statement is non-trivial and requires a proof using a theorem such as the FKG-inequality. We think Dubhashi et al.\ developed their proof to complete it.}.
%However Dubhashi presented a correct proof later using the FKG inequality~\cite[Th. 10]{dubhashi1996}.}

### 2025-06-28 paper/sec/intro.tex
Paper: References to supplementary material.   - Instead of referring to directories in the anonymized zip file, we refer to the AFP entry and github repo.   - Improve related work section regarding distinct-elements algorithms from Bar-Yossed and Alon et al.   - Add license files (use BSD consistent with the license of the AFP entries.)   - Add readme for the CVM_Transform session which is going to be references by swhids.

The supplementary material contains:
\begin{itemize}
\item formalization of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) using functional probabilistic invariants;
\item formalization of a library for negative association; and %in the directory Negative{\textunderscore}Association, which we intend to submit to the AFP as a separate contribution.
\item formalization of the CVM algorithm following the proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
\end{itemize}

>>>>

The supplementary material contains:
\begin{itemize}
\item formalization~\cite{CVM_Distinct_Elements-AFP} of the CVM algorithm, both the original version (\cref{alg:cvm}) and our new version (\cref{alg:cvm_new}) using functional probabilistic invariants;
\item formalization~\cite{Negative_Association-AFP} of a library for negative association; and %in the directory Negative{\textunderscore}Association, which we intend to submit to the AFP as a separate contribution.
\item formalization~\cite{CVM_Transforms-Github} of the CVM algorithm following the proof by Chakraborty et al.~\cite{chakraborty2022,chakraborty2023}.
\end{itemize}

### 2025-06-28 paper/sec/original_proof.tex
wip
Paper: References to supplementary material.   - Instead of referring to directories in the anonymized zip file, we refer to the AFP entry and github repo.   - Improve related work section regarding distinct-elements algorithms from Bar-Yossed and Alon et al.   - Add license files (use BSD consistent with the license of the AFP entries.)   - Add readme for the CVM_Transform session which is going to be references by swhids.

\section{Transformation-Based Proof}\label{sec:transformation_based_proof}
Here, we describe the transformation-based proof by Chakraborty et al.~\cite{chakraborty2023}, focusing on the challenging parts in its formalization.
An overview of the proof is shown in~\cref{fig:proof_overview}, which highlights, in part, why the transformation-based approach required more work to formalize, and why we developed our new approach using probabilistic invariants.
\begin{note}
In our supplementary material, the transformation-based formalization of CVM is in the directory \verb|CVM_Transforms|.
To formalize probabilistic transformations (relating two distributions), we built on an existing relational program logic in Isabelle/HOL~\cite{lochbihler2016}.
The formalization took \locold~lines which is considerably longer than the proof using our invariant-based technique (\locnew~lines).
\lipicsEnd\end{note}

@@

\subsection{A Bridging Transformation}
\begin{algorithm}[t]
	\caption{Modified CVM algorithm with independent coin flips. The function last{\textunderscore}index returns the index of the last occurrence of an element in the sequence, before the current iteration.}\label{alg:cvm_simul}
	\begin{algorithmic}[1]
  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ s.t. $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
  \State $\chi \gets \{\}, k \gets 0, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
  \State $b[i,j] \getsr \Ber(1/2)$ for $i,j \in \{1,\cdots,l\}$ \Comment perform $l^2$ unbiased independent coin flips
  \For{$i \gets 1$ to $l$}
    \If{$b[i,1]=b[i,2]=\cdots=b[i,k]=1$} \Comment insert $a_i$ if first $k$ flips are $1$s.
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$} \Comment if buffer $\chi$ is full
      \State $\chi \gets \{a \in \chi \;|\; b[\textrm{last{\textunderscore}index}(a),k+1] = 1\}$ \Comment keep elems.~whose $k+1$-th flip is $1$
      \State $k \gets k+1$
    \EndIf
    %\If{$|\chi| = n$}
    %  \State \Return $\bot$
    %\EndIf
  \EndFor
  \State \Return $2^k |\chi|$ \Comment estimate cardinality of $A$
  \end{algorithmic}
\end{algorithm}

>>>>

\section{Transformation-Based Proof}\label{sec:transformation_based_proof}
Here, we describe the transformation-based proof by Chakraborty et al.~\cite{chakraborty2023}, focusing on the challenging parts in its formalization.
An overview of the proof is shown in~\cref{fig:proof_overview}, which highlights, in part, why the transformation-based approach required more work to formalize, and why we developed our new approach using probabilistic invariants.
\begin{note}
The transformation-based formalization of the CVM algorthm is included in the supplementary material~\cite{CVM_Transforms-Github}.
To formalize probabilistic transformations (relating two distributions), we built on an existing relational program logic in Isabelle/HOL~\cite{lochbihler2016}.
The formalization took \locold~lines which is considerably longer than the proof using our invariant-based technique (\locnew~lines).
\lipicsEnd\end{note}

@@

\subsection{A Bridging Transformation}
\begin{algorithm}[t]
	\caption{Modified CVM algorithm with independent coin flips. The function last{\textunderscore}index returns the index of the last occurrence of an element in the sequence, before the current iteration.}\label{alg:cvm_simul}
	\begin{algorithmic}[1]
  \Require Stream elements $a_1,\dots,a_l$, $0 < \varepsilon$, $0 < \delta < 1$.
  \Ensure A cardinality estimate $R$ for set $A = \{ a_1,\dots,a_l \}$ such that $\prob \left( |R - |A| | > \varepsilon |A| \right) \leq \delta$
  \State $\chi \gets \{\}, k \gets 0, n = \ceil*{\frac{12}{\varepsilon^2} \ln{(\frac{6l}{\delta})} }$
  \State $b[i,j] \getsr \Ber(1/2)$ for $i,j \in \{1,\cdots,l\}$ \Comment perform $l^2$ unbiased independent coin flips
  \For{$i \gets 1$ to $l$}
    \If{$b[i,1]=b[i,2]=\cdots=b[i,k]=1$} \Comment insert $a_i$ if first $k$ flips are $1$s.
      \State $\chi \gets \chi \cup \{a_i\}$
    \Else
      \State $\chi \gets \chi - \{a_i\}$
    \EndIf
    \If{$|\chi| = n$} \Comment if buffer $\chi$ is full
      \State $\chi \gets \{a \in \chi \;|\; b[\textrm{last{\textunderscore}index}(a),k+1] = 1\}$ \Comment keep elems.~whose $k+1$-th flip is $1$
      \State $k \gets k+1$
    \EndIf
    %\If{$|\chi| = n$}
    %  \State \Return $\bot$
    %\EndIf
  \EndFor
  \State \Return $2^k |\chi|$ \Comment estimate cardinality of $A$
  \end{algorithmic}
\end{algorithm}

### 2025-06-28 paper/sec/related_work.tex
Paper: References to supplementary material.   - Instead of referring to directories in the anonymized zip file, we refer to the AFP entry and github repo.   - Improve related work section regarding distinct-elements algorithms from Bar-Yossed and Alon et al.   - Add license files (use BSD consistent with the license of the AFP entries.)   - Add readme for the CVM_Transform session which is going to be references by swhids.

\section{Related Work}\label{sec:related_work}
\subsection{Algorithms for the Distinct Elements Problem}
It is important to note that there are several practical solutions for the distinct elements problem.
The first solution was presented by Flajolet~\cite{flajolet1985} in 1985; however, like many other authors~\cite{flajolet2007,heule2013,pettie2021}, his solution makes the assumption that a fixed hash function can be regarded as a fully random function.
Alon et al.~\cite[Section 2.3]{alon1999} presented an easy remedy, which does not require such unmotivated model assumptions.
Their algorithm just relies on keeping track of the maximum of the hash values of the stream elements, where the hash function must be chosen uniformly from a pairwise independent family; the space complexity of this algorithm is $\bigo(\varepsilon^{-2} \ln (\delta^{-1}) b)$, where we assume that $b$ is the number of bits required to represent the stream elements.
This is slightly better than the CVM algorithm which requires $\bigo(\varepsilon^{-2} \ln (\delta^{-1}l) b)$ since there is no logarithmic dependency on $l$.

Later, Bar-Yossef et al.~\cite{baryossef2002}, Kane et al.~\cite{kane2010} and B\l{}asiok in 2020~\cite{blasiok2020} presented increasingly sophisticated solutions.
The last one by B\l{}asiok is optimal, with a space complexity of $\bigo(\varepsilon^{-2} \ln (\delta^{-1}) + b)$.
We~\cite{karayel2023} presented a version of the latter that preserves monotonicity and supports the merge-operation, which enables its use in distributed settings, such as Map-Reduce pipelines~\cite{dean2010}.
It should be noted that these recent algorithms are mostly of theoretical interest, as the constants, as well as the implementation complexity, are rather large.
What makes the CVM algorithm unique is its simplicity and the fact that it does not rely on hashing, which may enable more general use-cases than the traditional algorithms.

>>>>

\section{Related Work}\label{sec:related_work}
\subsection{Algorithms for the Distinct Elements Problem}
It is important to note that there are several practical solutions for the distinct elements problem.
The first solution was presented by Flajolet~\cite{flajolet1985} in 1985; however, like many other authors~\cite{flajolet2007,heule2013,pettie2021}, his solution makes the assumption that a fixed hash function can be regarded as a fully random function.
Alon et al.~\cite[Section 2.3]{alon1999} presented an easy remedy, which does not require such unmotivated model assumptions.
Their algorithm just relies on keeping track of the maximum of the hash values of the stream elements, where the hash function must be chosen uniformly from a pairwise independent family.

Later, Bar-Yossef et al.~\cite{baryossef2002}, Kane et al.~\cite{kane2010} and B\l{}asiok in 2020~\cite{blasiok2020} improved on the solution by Alon et al.
For example, Bar-Yossef et al.\ present a solution (Algorithm~3 in their work) with a space-complexity of $\bigo(\ln (\delta^{-1}) (\varepsilon^{-2}(\ln(\varepsilon^{-1})+\ln b) + b)))$, which can be implemented in practice.
This is slightly better than the CVM algorithm which requires $\bigo(\varepsilon^{-2} \ln (\delta^{-1}l) b)$. In particular, there is no dependency on the length of the stream $l$.
The more recent and more sophisticated solution by B\l{}asiok is space-optimal, with a space complexity of $\bigo(\varepsilon^{-2} \ln (\delta^{-1}) + b)$.
We~\cite{karayel2023} presented a version of the latter that preserves monotonicity and supports the merge-operation, which enables its use in distributed settings, such as Map-Reduce pipelines~\cite{dean2010}.
It should be noted that these recent algorithms are mostly of theoretical interest, as the constants, as well as the implementation complexity, are rather large.
What makes the CVM algorithm unique is its simplicity and the fact that it does not rely on hashing, which may enable more general use-cases than the traditional algorithms.
